Attrition and Achievement Gaps in Online Learning

### 2015-01-25 paper_v3.tex
final version

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A low response rate was expected for this subpopulation that was defined by low engagement. And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate). Missing values were multiply imputated by predictive mean matching using responses to all survey questions. All estimates were pooled across ten imputations, with standard errors adjusted for added variance. The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelor's or more advanced degree.

@@

Two research assistants independently developed codebooks for the resulting 448 non-empty open responses. Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations. The codebook was updated in each iteration to reliably fit the open response data. Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement. In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification. In the second iteration, the remaining responses were coded in the same way. In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see http://kizilcec.com/rsc/las2015attrition.pdf.} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.

Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item on the extent to which lower-priority distractions have hindered the learner's course progress; $M=2.73, SD=1.29$).

@@

The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (a sign of lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$). Reasons for learner attrition were measured using iteratively coded open responses. Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it. This list of reasons maps partially onto the factors identified in Study 1 (general time issues, difficulty, format \& content, and goals \& expectations). A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$). Hence, if a learner had not enough time, this tended to be the primary challenge.

To test hypotheses $H4ab$, we examined the 336 open responses that indicated having too little time. It revealed that no particular commitment was specified in about half of them (53\%). This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific resposnes. We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} While the list of reasons is certainly not collectively exhaustive, it was based on frequently mentioned terms in the open responses. Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported significantly lower levels of volitional control, i.e. their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion appeared to be hindered by low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.

The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$). A learner needs to be highly self-regulated to be successful in open learning environment that provide little guidance, such as MOOCs. Very few learners explicitly indicated issues with self-regulation in their open responses. They may not possess the metacognitive skills to consciously notice the issue or they preferred not to admit to this undesirable trait. Learners who reported a shortage of time without providing specific reasons were more likely to also indicate insufficient volitional control ($H4b$). This could be a useful proxy for identifying issues of self-regulation in future work. Overall, the results suggest that additional support and guidance to build stronger self-regulatory skills could help many learners persist in the course for longer.

@@

This study of over 100,000 online learners in 21 courses, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments. The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners of different gender and learners from different geographic locations; \emph{second}, the relation between self-ascribed success in online learning and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which appeared frequently related to learners' low capacity for volitional control. Moreover, we developed a survey item for measuring reasons for disengaging from a course (cf. Table \ref{tab:s2reas}), which should be tested and potentially extended based on more empirical evidence from additional courses.

In addressing research questions and testing hypotheses which emerged from a review of the literature, we sought to advance theory on attrition and provide empirical evidence. In line with early models of attrition that emphasized the importance of goal-directed behavior, we found goal striving to be higher in subjectively successful than unsuccessful learners. Moreover, volitional control was identified as a major contributor in reported time-related challenges with taking a MOOC. Tinto's \cite{tinto1975dropout} classic model of attrition emphasizes the role of academic and social integration, and incorporates individual learner characteristics as contributors to learners' decision to persist in a course of study. The current work provides evidence for the critical role of individual characteristics, such as gender, age, intentions, prior experiences, and geographic location. Social integration also remains an important factor, but in terms of perceived social belonging \cite{walton2007question}. In contrast to Tinto's concept of social and academic integration, which characterizes a general state in an institution, social belonging is a subjective experience based on perceptions and interpretations of environmental cues. Yet, this subjective experience is typically shared by individuals with a common identity, such as female learners in male-dominated fields. Models of attrition should be updated to account for the potential impact of psychological barriers, which can cause achievement gaps that would otherwise remain unnoticed or be misinterpreted.

@@

\section{Acknowledgments}
We thank Candace Thille, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for their assistance developing a codebook. The authors contributed equally to this work.

>>>>

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A low response rate was expected for this subpopulation that was defined by low engagement. And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate). Missing values were multiply imputed by predictive mean matching using responses to all survey questions. All estimates were pooled across ten imputations, with standard errors adjusted for added variance. The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelor's or more advanced degree.

@@

Two research assistants independently developed codebooks for the resulting 448 non-empty open responses. Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations. The codebook was updated in each iteration to reliably fit the open response data. Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement. In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification. In the second iteration, the remaining responses were coded in the same way. In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers). The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.

Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and the degree of volitional control they exerted (one item on the extent to which lower-priority distractions have hindered the learner's course progress; $M=2.73, SD=1.29$).

@@

The majority of survey respondents (550 of 756; 73\%) reported being at most `somewhat satisfied' with their progress in the course. While many of them (66\%) were substantially held back by other commitments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (a sign of lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$). Reasons for learner attrition were measured using iteratively coded open responses. Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it. This list of reasons maps partially onto the factors identified in Study 1 (general time issues, difficulty, format \& content, and goals \& expectations). A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$). Hence, if a learner had not enough time, this tended to be the primary challenge.

To test hypotheses $H4ab$, we examined the 336 open responses that indicated having too little time. It revealed that no particular commitment was specified in about half of them (53\%). This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific responses. We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} While the list of reasons is certainly not collectively exhaustive, it was based on frequently mentioned terms in the open responses. Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported significantly lower levels of volitional control, i.e. their progress was more severely hindered by lower-priority distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion appeared to be hindered by low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.

The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$). A learner needs to be highly self-regulated to be successful in open learning environment that provide little guidance, such as MOOCs. Very few learners explicitly indicated issues with self-regulation in their open responses. They may not possess the metacognitive skills to consciously notice the issue or they preferred not to admit to this undesirable trait. Learners who reported a shortage of time without providing specific reasons were more likely to admit getting distracted by lower-priority tasks, hinting at insufficient volitional control ($H4b$). This could be a useful proxy for identifying issues of self-regulation in future work. Overall, the results suggest that additional support and guidance to build stronger self-regulatory skills could help many learners persist in the course for longer.

@@

This study of over 100,000 online learners in 21 courses, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments. The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners of different gender and learners from different geographic locations; \emph{second}, the relation between self-ascribed success in online learning and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, most importantly having not enough time, which appeared frequently related to low levels of volitional control. Moreover, we developed a survey item for measuring reasons for disengaging from a course (cf. Table \ref{tab:s2reas}), which should be tested and potentially extended based on more empirical evidence from additional courses.

In addressing research questions and testing hypotheses which emerged from a review of the literature, we sought to advance theory on attrition and provide empirical evidence. In line with early models of attrition that emphasized the importance of goal-directed behavior, we found goal striving to be higher in subjectively successful than unsuccessful learners. Moreover, volitional control was identified as a major contributor to reported time-related challenges with taking a MOOC. Tinto's \cite{tinto1975dropout} classic model of attrition emphasizes the role of academic and social integration, and incorporates individual learner characteristics as contributors to learners' decision to persist in a course of study. The current work provides evidence for the critical role of individual characteristics, such as gender, age, intentions, prior experiences, and geographic location. Social integration also remains an important factor, but in terms of perceived social belonging \cite{walton2007question}. In contrast to Tinto's concept of social and academic integration, which characterizes a general state in an institution, social belonging is a subjective experience based on perceptions and interpretations of environmental cues. Yet, this subjective experience is typically shared by individuals with a common identity, such as female learners in male-dominated fields. Models of attrition should be updated to account for the potential impact of psychological barriers, which can cause achievement gaps that would otherwise remain unnoticed or be misinterpreted.

@@

\section{Acknowledgments}
We thank Candace Thille, Octavia Zahrt, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for their assistance developing a codebook. The authors contributed equally to this work.
